Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-28 paper/paper.tex
space saved

As   demonstrated  by   the  gap   between  the   \RGCLine\  and   the
\UseLine\  lines, a  large number  of cells  which are  unused by  the
program are  still copied during  RGC. LGC does  a much better  job of
closing this  gap but still falls  short of the precision  achieved by
LGC in case of eager  languages ~\cite{asati14lgc}.  A major source of
inefficiency  in LGC  is multiple  traversals of  already copied  heap
cells.  Since LGC does not mark  the heap cells after the first visit,
the  same cells  can  be repeatedly  visited  with different  liveness
states.  This can  be  seen in  the cells touched  per  GC for  $huffman$
benchmark in Table~\ref{fig:memory-usage}, even though it uses only 72
cells, number of cells touched is  88.  We have mitigated this problem
by adding checks to avoid  multiple traversals of closure and function
arguments.

@@

Our proofs will be for a single round of minefield execution i.e., the
evaluation  of  $(\mainpgm)$  to  its  WHNF  driven  by  the  printing
mechanism~(Section \ref{sec:semantics}).   With minor  variations, the
proof will also be applicable  for subsequent evaluations initiated by
$\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.
%
\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
minefield execution of a program.  We  show by induction on the number $n$
of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
transition cannot  enter a  \bang\ state.  Since  our programs  are in
ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
does not involve dereferencing, and thus cannot result in a \bang.

@@

For the  {\sc car-1-clo}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell,  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  GC with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus,   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}

@@

\begin{proof}
Assume to  the contrary that  a program $P$  enters the
\bang\  state.  We can  transform  $P$  to a  call-less
program $P'$ such that  the minefield executions of $P$
and $P'$  are identical  except for change  of variable
names.             However,                 by
Lemma~\ref{lemma:call-less-cannot-go-bang}, we know that
$P'$,   a   call-less   program,   cannot   enter   the
\bang\  state.  Therefore,   $P$  also  cannot  enter  the
\bang\ state.
\end{proof}

@@

After the  TM has  halted, there  are further  ``cleanup'' derivations
that derive  $\epsilon$ only if  the invariant has been  maintained so
far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
and                $\nont{\mathit{final}}{c}               \rightarrow
\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
\acdr\ on the right of the tape head.  This completes the
reduction.\!
\end{proof}

@@

\begin{figure}[t]
  \begin{minipage}{.45\columnwidth}
    \begin{align*}
      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
      \acdr\nont{\mathit{all}}{}
    \end{align*}
  \end{minipage}
  \begin{minipage}{.45\columnwidth}
    \input{equivalent-program}
  \end{minipage}\\
  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
  \caption{A possible grammar generated by the proof of
    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
  grammar.}\label{fig:turing-pgm}
\vspace*{-2mm}
\end{figure}

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
sentential form by $\epsilon$. However, if  the TM does not accept the
input string, then every sentential form derived from the start symbol
would    have     at    least    one    non-terminal from the grammar
of Lemma~\ref{lemma:grammar-undecidable} that is different
from $\nont{\mathit{final}}{c}$.   Thus,  $\$$  would  not  be
derivable from the grammar.
\end{proof}

@@

%%\clearpage
\balance
\bibliography{fun_hra}{}
\bibliographystyle{abbrv}
\end{document}

>>>>

As   demonstrated  by   the  gap   between  the   \RGCLine\  and   the
\UseLine\  lines, a  large number  of cells  which are  unused by  the
program are  still copied during  RGC. LGC does  a much better  job of
closing this  gap but still falls  short of the precision  achieved by
LGC in case of eager  languages ~\cite{asati14lgc}.  A major source of
inefficiency  in LGC  is multiple  traversals of  already copied  heap
cells.  Since LGC does not mark  the heap cells after the first visit,
the  same cells  can  be repeatedly  visited  with different  liveness
states.   This  can   be  seen  in  the  cells  touched   per  GC  for
\verb@huffman@  benchmark in  Table~\ref{fig:memory-usage}.  Even  though the  benchmark  uses  72 cells  maximum,  the
average  number  of cells  touched per GC is 88. During  a  GC,  for every  traversal  the
arguments of a  closure are garbage collected using  the same automata
(stored in  the closure itself). We mark the arguments {\em
  visited}  during  the  first  traversal  and  avoid subsequent
traversals.

@@

Our proofs will be for a single round of minefield execution i.e., the
evaluation  of  $(\mainpgm)$  to  its  WHNF  driven  by  the  printing
mechanism~(Section \ref{sec:semantics}).   With minor  variations, the
proof will also be applicable  for subsequent evaluations initiated by
$\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.
%
\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
minefield execution of a program.  We  show by induction on the number $n$
of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, e_{\mainpgm})$.  A call  to $GC(\ldots)$ in  this state
does  nothing. Also,  the  $\rightsquigarrow$
transition cannot  enter a  \bang\ state.  This is because  our programs  are in
ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
does not involve dereferencing, and thus cannot result in a \bang.

@@

For the  {\sc car-1-clo}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell,  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  GC with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus,   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof} \vspace*{-1mm}

@@

\begin{proof}
Assume to  the contrary that  a program $P$  enters the
\bang\  state.  We can  transform  $P$  to a  call-less
program $P'$ such that  the minefield executions of $P$
and $P'$  are identical  except for change  of variable
names.             However,                 by
Lemma~\ref{lemma:call-less-cannot-go-bang}, we know that
$P'$,   a   call-less   program,   cannot   enter   the
\bang\  state.  Therefore,   $P$  also  cannot  enter  the
\bang\ state.
\end{proof} \vspace*{-1mm}

@@

After the  TM has  halted, there  are further  ``cleanup'' derivations
that derive  $\epsilon$ only if  the invariant has been  maintained so
far.    For   every   symbol   $c$,  we   introduce   a   non-terminal
$\nont{\mathit{final}}{c}$  where   $\nont{\mathit{final}}{}$  is  the
final state  of the  TM. We add  productions $\nont{\mathit{final}}{c}
\rightarrow         \acar        \nont{\mathit{final}}{c}$         and
$\nont{\mathit{final}}{c} \rightarrow  \acdr \nont{\mathit{final}}{c}$
for cleaning up the \bcar\ and \bcdr\  symbols on the left of the head
and                $\nont{\mathit{final}}{c}               \rightarrow
\nont{\mathit{final}}{c}\bcar$      and      $\nont{\mathit{final}}{c}
\rightarrow \nont{\mathit{final}}{c}\bcdr$ for  cleaning up \acar\ and
\acdr\ on the right of the tape head.  This completes the
reduction.\!
\end{proof} \vspace*{-2mm}

@@

\begin{figure}[t]
  \begin{minipage}{.45\columnwidth}
    \begin{align*}
      \nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{i}{\acar} &\rightarrow \clazy\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
      \nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
      \acdr\nont{\mathit{all}}{}
    \end{align*}
  \end{minipage}
  \begin{minipage}{.45\columnwidth}
    \input{equivalent-program}
  \end{minipage}\\
  \mbox{}\hfill(a)\hfill(b)\hfill\mbox{}
  \caption{A possible grammar generated by the proof of
    Lemma~\ref{lemma:grammar-undecidable} and a program to realize the
  grammar.}\label{fig:turing-pgm}
\vspace*{-4mm}
\end{figure}

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every   time   replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the
sentential form by $\epsilon$. However, if  the TM does not accept the
input string, then every sentential form derived from the start symbol
would    have     at    least    one    non-terminal from the grammar
of Lemma~\ref{lemma:grammar-undecidable} that is different
from $\nont{\mathit{final}}{c}$.   Thus,  $\$$  would  not  be
derivable from the grammar.
\end{proof} \vspace*{-2mm}

@@

\clearpage
\balance
\bibliography{fun_hra}{}
\bibliographystyle{abbrv}
\end{document}
